\documentclass[11pt,a4paper]{article}

% --- Packages ---
\usepackage[margin=1in]{geometry}
\usepackage{amsmath,amssymb,amsfonts}
\usepackage{booktabs}
\usepackage{graphicx}
\usepackage{hyperref}
\usepackage{xcolor}
\usepackage{colortbl}
\usepackage{caption}
\usepackage{subcaption}
\usepackage{pgfplots}
\usepackage{pgfplotstable}
\pgfplotsset{compat=1.18}
\usepgfplotslibrary{colorbrewer}

% Color definitions
\definecolor{apn-none}{HTML}{E64B35}
\definecolor{apn-tanh}{HTML}{F39B7F}
\definecolor{apn-ffn}{HTML}{B83226}
\definecolor{apan-rmsnorm}{HTML}{7B2D8B}
\definecolor{apan-tanh}{HTML}{C77CFF}
\definecolor{deltanet}{HTML}{4DBBD5}
\definecolor{deltanet-pm}{HTML}{91D1C2}
\definecolor{deltanet-ffn}{HTML}{00868B}
\definecolor{transformer}{HTML}{3C5488}
\definecolor{transformer-pm}{HTML}{8491B4}
\definecolor{gfr-k1}{HTML}{7E6148}
\definecolor{gfr-k4}{HTML}{B09C85}
\definecolor{gfr-ffn}{HTML}{DC9C6B}

\hypersetup{colorlinks=true, linkcolor=blue, citecolor=blue, urlcolor=blue}

\title{Seq-CIFAR-10 Benchmark Report:\\
Comparing APAN, APN, GFR, DeltaNet, and Transformer Architectures}
\author{Zhixin Lu}
\date{April 2026}

\begin{document}
\maketitle

% ============================================================================
\begin{abstract}
We benchmark five sequence model architectures on the sequential CIFAR-10 task:
the \textbf{APAN} (Associative Plasticity Additive Networks), the original Hebbian memory model using an additive (attractor) update;
the \textbf{APN} (Associative Plasticity Network), a prediction-error variant mapped onto the gated delta-rule Triton kernel;
the \textbf{GFR} (Gated Forgetting Recurrence), a simplified element-wise gated scan model;
\textbf{DeltaNet}, a delta-rule linear attention model; and a standard \textbf{softmax Transformer} with FlashAttention-2.
All experiments use 10 layers, run for 200 epochs on a single NVIDIA H200 GPU, and share the same outer architecture.

Key findings:
(1)~\textbf{APAN with RMSNorm keys (64.2\%, 105K params) is the best model at the $\sim$103K parameter budget}---outperforming prediction-error APN (58.3\%), DeltaNet (60.1\%), the Transformer (59.8\%), and all GFR variants---while training $4\times$ faster than DeltaNet.
(2)~The original APAN's Hebbian attractor update requires careful stability analysis; RMSNorm key normalization ensures $\|k\|^2 = d$, and tanh-parameterized $\eta \in (-1,1)$ guarantees unconditional stability.
(3)~APAN is implemented on the \emph{unmodified} delta-rule chunk kernel by negating both $v$ and $\beta$---no kernel changes required.
(4)~The prediction-error APN (58.3\%) is uniquely resistant to overfitting (0.17-point drop) but has a hard accuracy ceiling.
(5)~FFN augmentation hurts both APAN and APN (APAN+FFN: 57.9\% vs.\ APAN: 64.2\%; APN+FFN: 56.8\% vs.\ APN: 58.3\%), while helping GFR and DeltaNet.
(6)~Multi-head APN ($H{=}4$) diverges to NaN due to unconstrained $\eta$; APAN's tanh-bounded $\eta$ prevents this failure mode.
\end{abstract}

% ============================================================================
\section{Introduction}

The Associative Plasticity Additive Networks (APAN) is a biologically-inspired recurrent model that maintains a Hebbian plastic memory matrix $M_t$, updated at every time-step using the layer's full output (the \emph{additive} signal $\mathbf{v} + S^\top\mathbf{k}$) as the learning signal.
The prediction-error variant (APN) replaces the Hebbian signal with a ``surprise'' signal.
By mapping both recurrences onto the \emph{gated delta-rule} chunk kernel from the Flash Linear Attention (FLA) library, we achieve efficient GPU training with $O(TD)$ complexity instead of the $O(T^2 D)$ cost of standard softmax attention.

This report summarizes all 30 experiments conducted on the \textbf{Seq-CIFAR-10} benchmark, where each $32 \times 32$ CIFAR-10 image is flattened into a sequence of $T = 1024$ RGB pixels ($3$-dimensional features), and the model classifies from the last time-step.
We compare APAN (original Hebbian), APN (prediction-error), GFR, DeltaNet, and the Transformer in single-head and multi-head configurations, with and without FFN blocks, across parameter-matched groups.

% ============================================================================
\section{Model Architectures}

All models share the same outer wrapper:
\[
\texttt{Linear}(3 \to D) \;\to\; \tanh \;\to\; \underbrace{[\texttt{Layer}_i + \text{residual}]}_{i=1}^{L} \;\to\; \texttt{LayerNorm} \;\to\; \texttt{Linear}(D \to 10)
\]

\subsection{APAN Layer (Associative Plasticity Additive Networks)}
\label{sec:apan}

The APAN implements the ``$W + M$'' principle in its purest Hebbian form.
Each layer's effective weight is $W + S_t$, where $W$ is a static weight trained by backpropagation and $S_t$ is a plastic state that adapts online within the forward pass.
The layer's output \emph{is} the retrieval from the combined weight:

\paragraph{The $W + M$ Story.}
Given input $\mathbf{x}_t$, define $\mathbf{k}_t = \text{norm}(\mathbf{x}_t)$ (the normalized key/pre-synaptic activity).
The output is:
\begin{equation}
\mathbf{h}_t = (W + S_t)^\top \mathbf{k}_t = \underbrace{W^\top \mathbf{k}_t}_{\mathbf{v}_t\;\text{(static)}} + \underbrace{S_t^\top \mathbf{k}_t}_{\text{plastic recall}}
\end{equation}
This is the ``$W + M$'' retrieval: the static weight $W$ provides innate/trained knowledge, while the plastic matrix $S_t$ accumulates experience within the sequence.

\paragraph{Hebbian plasticity rule.}
The plastic state $S_t \in \mathbb{R}^{d \times d}$ (per head) evolves via:
\begin{equation}\label{eq:apan-update}
S_t = \lambda \, S_{t-1} + \frac{\eta(1-\lambda)}{d} \, \mathbf{k}_t \otimes \underbrace{(\mathbf{v}_t + S_{t-1}^\top \mathbf{k}_t)}_{\mathbf{h}_t\;\text{(full output = learning signal)}}
\end{equation}
The learning signal is $\mathbf{h}_t = \mathbf{v}_t + S_{t-1}^\top \mathbf{k}_t$---the layer's own output.
This creates a Hebbian \emph{attractor}: the memory reinforces associations that it already partially recalls, deepening basins of attraction for stored patterns.

\paragraph{Stability analysis.}
The effective per-step gain is:
\begin{equation}
\alpha = \lambda + \frac{\eta(1-\lambda)\|\mathbf{k}_t\|^2}{d}
\end{equation}
For stability we require $|\alpha| < 1$. The positive-feedback nature of the Hebbian update means this must be explicitly enforced.

\paragraph{Key normalization for stability.}
We normalize keys so that $\|\mathbf{k}\|^2$ is bounded:
\begin{itemize}
\item \textbf{RMSNorm} (recommended): $\mathbf{k} = \text{RMSNorm}(\mathbf{x})$ ensures $\|\mathbf{k}\|^2 = d$ exactly, giving $\alpha = \lambda + \eta(1-\lambda)$.
\item \textbf{L2-unit}: $\mathbf{k} = \mathbf{x}/\|\mathbf{x}\|$ gives $\|\mathbf{k}\|^2 = 1$, hence $\alpha = \lambda + \eta(1-\lambda)/d$ (very loose constraint on $\eta$).
\item \textbf{tanh}: $\mathbf{k} = \tanh(\mathbf{x})$ gives $\|\mathbf{k}\|^2 \le d$, same bound as RMSNorm.
\end{itemize}

\paragraph{Tanh parameterization of $\eta$.}
We parameterize $\eta = \tanh(\hat\eta)$ where $\hat\eta$ is the raw learnable parameter. This gives $\eta \in (-1, 1)$ with smooth gradients everywhere.
With RMSNorm keys ($\|\mathbf{k}\|^2 = d$) and $\lambda = 0.99$:
\begin{equation}
\alpha = \lambda + \eta(1-\lambda) = 0.99 + \eta \cdot 0.01 \in (0.98, 1.0)
\end{equation}
This is \textbf{unconditionally stable} ($|\alpha| < 1$), \textbf{always positive} (no oscillation), and supports both Hebbian ($\eta > 0$) and anti-Hebbian ($\eta < 0$) learning.

\paragraph{Implementation via the delta-rule kernel (no kernel modification).}
The FLA chunk kernel computes the \emph{delta rule}:
\begin{equation}
S_t = e^g \, S_{t-1} + \beta_{\text{kern}} \, \mathbf{k}_t \otimes (\mathbf{v}_{\text{kern}} - S_{t-1}^\top \mathbf{k}_t)
\end{equation}
We need the Hebbian form $S_t = \lambda\, S_{t-1} + \beta\, \mathbf{k}_t \otimes (\mathbf{v} + S_{t-1}^\top \mathbf{k}_t)$ (note the $+$ sign).
The key insight: \textbf{negate both $\mathbf{v}$ and $\beta$} before passing to the kernel:
\begin{align}
\text{Kernel input:} \quad &v_{\text{kern}} = -\mathbf{v}, \quad \beta_{\text{kern}} = -\beta \\
\text{Kernel computes:} \quad &S_t = e^g\, S_{t-1} + (-\beta)\, \mathbf{k} \otimes ((-\mathbf{v}) - S^\top \mathbf{k}) \\
&= e^g\, S_{t-1} + \beta\, \mathbf{k} \otimes (\mathbf{v} + S^\top \mathbf{k}) \quad \checkmark
\end{align}
The minus signs cancel perfectly, recovering the Hebbian update with \textbf{zero modifications} to the highly-optimized Triton kernel.
The gate is set to $g = \log(\lambda)$ to recover the exponential decay.
The kernel output $o = q^\top S_t$ (with $q = k$) gives the plastic recall $S_t^\top \mathbf{k}_t$, and we add $\mathbf{v}_t$ to get the full output $\mathbf{h}_t$.

\textbf{Parameters per layer:} $D^2$ (one linear $W$) $+ D$ (RMSNorm) $+ 2D$ (output LayerNorm) $+ 2$ (scalars $\eta, \lambda$) $\approx D^2$.

% -----------------------------------------------------------------------

\subsection{APN Layer (Prediction-Error Variant)}
\label{sec:apn}

The APN (prediction-error variant) shares the same ``$W + M$'' output structure as APAN, but uses a different learning signal for the plasticity update.

\paragraph{Output.}
The layer output combines static and plastic pathways:
\begin{equation}
\mathbf{o}_t = \underbrace{W^\top \text{act}(\mathbf{x}_t)}_{\mathbf{v}_t\text{ (static)}} + \underbrace{S_t^\top \text{act}(\mathbf{x}_t)}_{\text{plastic recall}} = (W + S_t)^\top \text{act}(\mathbf{x}_t)
\end{equation}

\paragraph{Plasticity rule.}
The plastic state $S_t \in \mathbb{R}^{D \times D}$ decays and updates via:
\begin{equation}\label{eq:apn-update}
S_t = \lambda \, S_{t-1} + \frac{\eta(1-\lambda)}{D} \, \mathbf{k}_t \otimes \underbrace{(\mathbf{v}_t - S_{t-1}^\top \mathbf{k}_t)}_{\mathbf{h}_t\text{ (prediction error)}}
\end{equation}
where $\mathbf{k}_t = \text{act}(\mathbf{x}_t)$, $\mathbf{v}_t = W \cdot \text{act}(\mathbf{x}_t)$, $\lambda \in (0,1)$ is a trainable decay, and $\eta$ is a trainable learning rate scalar.

\paragraph{Distinction from APAN.}
While APAN (Section~\ref{sec:apan}) uses the full output $\mathbf{h}_t = \mathbf{v}_t + S_{t-1}^\top \mathbf{k}_t$ as the learning signal (Hebbian attractor),
APN uses the \emph{prediction error} $\mathbf{v}_t - S_{t-1}^\top \mathbf{k}_t$ (delta rule).
This provides negative feedback: once the plastic memory has learned an association ($S^\top \mathbf{k} \approx \mathbf{v}$), the update vanishes automatically, making the recurrence unconditionally stable for any $\eta > 0$ without requiring key normalization.
However, this stability comes at a cost: the prediction-error signal is weaker than the Hebbian attractor signal, limiting the model's ability to deeply encode associations.
As our experiments show, APAN (64.2\%) significantly outperforms APN (58.3\%) at the same parameter budget.
The ``$W + M$'' interpretation is preserved: $W$ encodes innate/trained knowledge, $S$ accumulates experience within a sequence, and the effective weight $(W + S_t)$ grows richer over time.

\textbf{Parameters per layer:} $D^2$ (one linear projection $W$) $+ 2D$ (LayerNorm) $+ 2$ (scalars $\eta, \lambda$) $\approx D^2$.

\subsection{GFR Layer (Gated Forgetting Recurrence)}

\paragraph{Intuition.}
GFR is a minimalist baseline that tests how much of APAN/APN's performance comes from its $D \times D$ matrix state (which can store associative key-value patterns) versus simple temporal memory.
Each neuron maintains $K$ independent scalar memory slots, each decaying exponentially---no cross-neuron interaction, no outer products, no associative storage.
If GFR approaches APAN/APN performance, then the matrix state is unnecessary; if there is a large gap, the associative capacity is proven essential.
GFR uses the HGRN kernel (element-wise gated prefix scan), which is trivially parallel and extremely fast.

\paragraph{Update rule.}
Each GFR layer maintains $K$ scalar memory slots per neuron (total state: $D \cdot K$ scalars):
\begin{equation}
m_{i,k,t} = \lambda_k \, m_{i,k,t-1} + z_{i,t}, \quad
h_{i,t} = z_{i,t} + \sum_{k=1}^{K} \eta_k \, m_{i,k,t}
\end{equation}
where $\mathbf{z}_t = W \cdot \text{act}(\mathbf{x}_t)$ is the projected input.

\textbf{Parameters per layer:} $D^2$ (projection $W$) $+ 2K$ (scalars $\lambda, \eta$) $+ 2D$ (LayerNorm) $\approx D^2$.

\subsection{DeltaNet Layer}

\paragraph{Intuition.}
DeltaNet is a \emph{linear attention} model that maintains an associative memory via the delta rule---the same error-correcting update used in classical Widrow--Hoff learning.
Unlike APAN/APN's single shared projection, DeltaNet decouples what to store ($V$), how to index it ($K$), and how to query it ($Q$) via separate linear projections, giving the model more representational flexibility at the cost of $3\times$ more parameters per layer.
The input-dependent step size $\beta_t = \sigma(\cdot) \in (0,1)$ allows the model to selectively update: large $\beta$ for informative tokens, small $\beta$ for noise.

\paragraph{Update rule.}
\begin{equation}
S_t = S_{t-1} + \beta_t \, \mathbf{k}_t \otimes \bigl(\mathbf{v}_t - S_{t-1}^\top \mathbf{k}_t\bigr), \quad \mathbf{o}_t = \mathbf{q}_t^\top S_t
\end{equation}
where $\beta_t = \sigma(W_\beta \mathbf{x}_t)$ is a learned, input-dependent step size.
The delta-rule error signal $\mathbf{v}_t - S_{t-1}^\top \mathbf{k}_t$ ensures the memory converges toward storing the correct key-value association and is unconditionally stable (same as APN).
DeltaNet uses the same \texttt{chunk\_gated\_delta\_rule} Triton kernel as APN/APAN, benefiting from the same GPU optimizations described in Section~\ref{sec:kernels}.

\textbf{Parameters per layer:} $3D^2$ (Q/K/V projections) $+ D$ ($\beta$ projection) $+ 2D$ (LayerNorm) $\approx 3D^2$.

\subsection{Transformer Layer}

\paragraph{Intuition.}
The Transformer uses \emph{softmax attention} to compute all-pairs token interactions, allowing each position to attend to any previous position with learned importance weights.
This gives maximal representational power but at $O(T^2)$ cost in sequence length.
We include it as the gold-standard upper bound: if a linear-time model can approach the Transformer's accuracy, it validates the efficiency of recurrent state-space approaches.
We use FlashAttention-2 for memory-efficient training, which avoids materializing the $T \times T$ attention matrix in HBM.

\paragraph{Architecture.}
A pre-norm causal Transformer block:
\[
\texttt{LN} \to \texttt{MHA} \to \text{residual} \to \texttt{LN} \to \texttt{FFN}(D \to 4D \to D) \to \text{residual}
\]
The FFN is \emph{integral} to the Transformer (unlike APAN/APN where it is optional and harmful); removing it would halve the model's capacity.

\textbf{Parameters per layer:} $4D^2$ (QKV + output proj) $+ 8D^2$ (FFN) $+ 4D$ (two LayerNorms) $\approx 12D^2$.

\subsection{Parameter Count Summary}

\begin{table}[h]
\centering
\caption{Parameters per layer and total model parameters ($L=10$).}
\label{tab:params}
\begin{tabular}{lccc}
\toprule
\textbf{Model} & \textbf{Per-layer} & \textbf{Total (10L)} & \textbf{Relative} \\
\midrule
\multicolumn{4}{l}{\textit{No FFN ($\approx$103K parameter-matched)}} \\
\midrule
APAN ($D=100$, RMSNorm) & $\approx D^2 = 10$K & 104,630 & $1\times$ \\
APN ($D=100$) & $\approx D^2 = 10$K & 103,630 & $1\times$ \\
GFR $K{=}1$ ($D=100$) & $\approx D^2 = 10$K & 103,630 & $1\times$ \\
GFR $K{=}4$ per-slot ($D=100$) & $\approx D^2 = 10$K & 103,690 & $1\times$ \\
DeltaNet ($D=58$) & $\approx 3D^2 = 10$K & 103,598 & $1\times$ \\
Transformer ($D=29$) & $\approx 12D^2 = 10$K & 102,554 & $1\times$ \\
\midrule
\multicolumn{4}{l}{\textit{With FFN ($\approx$1.2M parameter-matched)}} \\
\midrule
APN + FFN ($D=115$) & $\approx 9D^2$ & 1,196,720 & $11.5\times$ \\
GFR $K{=}1$ + FFN ($D=115$) & $\approx 9D^2$ & 1,196,720 & $11.5\times$ \\
DeltaNet + FFN ($D=104$) & $\approx 11D^2$ & 1,196,634 & $11.5\times$ \\
Transformer ($D=100$) & $\approx 12D^2$ & 1,205,610 & $11.6\times$ \\
\midrule
\multicolumn{4}{l}{\textit{Width-matched ($D=100$, no FFN)}} \\
\midrule
DeltaNet ($D=100$) & $\approx 3D^2 = 30$K & 304,610 & $2.9\times$ \\
\bottomrule
\end{tabular}
\end{table}

% ============================================================================
\section{Experimental Setup}

\begin{itemize}
\item \textbf{Hardware:} Single NVIDIA H200 GPU (141 GB HBM3e, 228 KB shared memory per SM)
\item \textbf{Task:} Seq-CIFAR-10 (1024-step sequences, 10-class classification)
\item \textbf{Optimizer:} AdamW ($\text{lr} = 10^{-3}$, weight decay $= 10^{-4}$, grad clip $= 1.0$)
\item \textbf{Schedule:} Cosine annealing with 20-epoch linear warmup (unless noted)
\item \textbf{Batch size:} 64
\item \textbf{Epochs:} 200
\item \textbf{Seed:} 42
\item \textbf{Triton version:} $\ge$3.3.0, $<$3.4.0 (pinned due to Hopper backward kernel bug in $\ge$3.4)
\end{itemize}

% ============================================================================
\section{Results}

\subsection{Complete Results Summary}

\begin{table}[h]
\centering
\caption{All 30 experiments (200 epochs unless noted). Bold: best per group. ``Drop'' = peak $-$ final test accuracy.}
\label{tab:results}
\resizebox{\textwidth}{!}{%
\begin{tabular}{llcccccccc}
\toprule
\textbf{Model} & \textbf{Config} & \textbf{$D$} & \textbf{$H$} & \textbf{Params} & \textbf{Peak Acc (\%)} & \textbf{Peak Ep} & \textbf{Final Acc (\%)} & \textbf{Drop} & \textbf{s/ep} \\
\midrule
\multicolumn{10}{l}{\textit{Group 1: $\approx$1.2M params (with FFN or large Transformer)}} \\
\midrule
Transformer & $H{=}4$, warmup20 & 100 & 4 & 1,205,610 & \textbf{65.16} & 183 & 65.13 & 0.03 & 85.1 \\
Transformer & $H{=}1$, warmup20 & 100 & 1 & 1,205,610 & 63.50 & 180 & 63.46 & 0.04 & 39.3 \\
DeltaNet+FFN & $H{=}4$, warmup20 & 104 & 4 & 1,199,754 & 63.56 & 145 & 63.39 & 0.17 & 59.1 \\
DeltaNet+FFN & $H{=}1$, warmup20 & 104 & 1 & 1,196,634 & 62.80 & 198 & 62.80 & 0.00 & 112.1 \\
GFR $K{=}1$+FFN & warmup20 & 115 & 1 & 1,196,720 & 58.93 & 192 & 58.75 & 0.18 & 33.3 \\
APN+FFN & $H{=}1$, warmup20 & 115 & 1 & 1,196,720 & 56.84 & 177 & 56.68 & 0.16 & 55.3 \\
APAN+FFN & RMSNorm keys & 100 & 1 & 906,630 & 57.89 & $\sim$195 & 57.86 & 0.03 & $\sim$35 \\
APAN+FFN & tanh keys & 100 & 1 & 905,630 & 52.12 & $\sim$100 & 50.44 & 1.68 & $\sim$35 \\
APN+FFN & $H{=}4$, per-head & 116 & 4 & 1,217,626 & 40.58$^\dagger$ & 9 & 10.00 & --- & 37.7 \\
\midrule
\multicolumn{10}{l}{\textit{Group 2: $\approx$103K params (no FFN, parameter-matched)}} \\
\midrule
\textbf{APAN} & \textbf{RMSNorm keys, act=none} & 100 & 1 & 104,630 & \textbf{64.22} & $\sim$190 & 63.96 & 0.26 & $\sim$25 \\
APAN & RMSNorm keys, $H{=}4$ & 100 & 4 & 104,690 & 62.28 & $\sim$155 & 58.09 & 4.19 & 28.3 \\
DeltaNet & param-match & 58 & 1 & 103,598 & 60.06 & 18 & 55.40 & 4.66 & 69.0 \\
Transformer & param-match, warmup20 & 29 & 1 & 102,554 & 59.75 & 38 & 55.89 & 3.86 & 32.6 \\
APN & act=none, warmup20 & 100 & 1 & 103,630 & 58.34 & 189 & 58.17 & 0.17 & 25.5 \\
APAN & tanh keys, no norm & 100 & 1 & 103,630 & 56.42 & $\sim$185 & 56.15 & 0.27 & $\sim$25 \\
APN & act=none, no warmup & 100 & 1 & 103,630 & 58.12 & 163 & 57.96 & 0.16 & 26.0 \\
GFR $K{=}1$+FFN & $D{=}34$, warmup20 & 34 & 1 & 105,974 & 57.57 & 39 & 51.89 & 5.68 & 18.9 \\
GFR $K{=}4$ & per-slot, warmup20 & 100 & 1 & 103,690 & 56.00 & 199 & 55.70 & 0.30 & 22.7 \\
GFR $K{=}4$ & per-neuron, warmup20 & 100 & 1 & 111,610 & 56.21 & 196 & 55.56 & 0.65 & 22.8 \\
GFR $K{=}1$ & per-slot, warmup20 & 100 & 1 & 103,630 & 53.57 & 199 & 53.01 & 0.56 & 14.0 \\
GFR $K{=}1$ & per-neuron, warmup20 & 100 & 1 & 105,610 & 52.43 & 189 & 51.73 & 0.70 & 14.0 \\
APN & act=tanh, better init & 100 & 1 & 103,630 & 53.15 & 168 & 52.50 & 0.65 & 80.1 \\
APN & act=tanh & 100 & 1 & 103,630 & 51.69 & 162 & 51.16 & 0.53 & 26.2 \\
APN & $H{=}4$, per-head & 100 & 4 & 103,690 & 10.00$^\dagger$ & --- & 10.00 & --- & 26.1 \\
\midrule
\multicolumn{10}{l}{\textit{Group 3: Width-matched ($D{=}100$, no FFN)}} \\
\midrule
APAN & RMSNorm keys & 100 & 1 & 104,630 & \textbf{64.22} & $\sim$190 & 63.96 & 0.26 & $\sim$25 \\
DeltaNet & no warmup & 100 & 1 & 304,610 & 62.28 & 11 & 60.80 & 1.48 & 100.4 \\
DeltaNet & warmup20 & 100 & 1 & 304,610 & 59.26 & 21 & 58.73 & 0.53 & 100.4 \\
APN & act=none, warmup20 & 100 & 1 & 103,630 & 58.34 & 189 & 58.17 & 0.17 & 25.5 \\
\bottomrule
\end{tabular}%
}

\smallskip
{\footnotesize ${}^\dagger$ Diverged to NaN during training (see Section~\ref{sec:nan}).}
\end{table}

% ============================================================================
\subsection{Key Finding 1: APAN Is the Most Parameter-Efficient Model}

\begin{figure}[h]
\centering
\begin{tikzpicture}
\begin{axis}[
    width=0.85\textwidth,
    height=0.55\textwidth,
    xlabel={Epoch},
    ylabel={Test Accuracy (\%)},
    xmin=0, xmax=200,
    ymin=35, ymax=66,
    grid=major,
    grid style={gray!30},
    legend style={at={(0.98,0.02)}, anchor=south east, font=\footnotesize, draw=none, fill=white, fill opacity=0.8},
    every axis plot/.append style={thick},
    title={\textbf{Parameter-matched group ($\approx$103K params, no FFN)}},
    title style={font=\normalsize},
]

\addplot[color=apan-rmsnorm, thick] table[x=epoch, y=test/acc, col sep=comma]
    {../report_data/apan_L10_D100_rmsnorm_noFFN.csv};
\addlegendentry{APAN RMSNorm (105K) --- \textbf{64.2\%}}

\addplot[color=apn-none, thick] table[x=epoch, y=test/acc, col sep=comma]
    {../report_data/apn_L10_D100_H200_B64_act-none_warmup20.csv};
\addlegendentry{APN act=none (103K) --- 58.3\%}

\addplot[color=apan-tanh, thick, dashed] table[x=epoch, y=test/acc, col sep=comma]
    {../report_data/apan_L10_D100_tanh_noFFN.csv};
\addlegendentry{APAN tanh (103K) --- 56.4\%}

\addplot[color=deltanet-pm, thick] table[x=epoch, y=test/acc, col sep=comma]
    {../report_data/deltanet_L10_D58_H200_B64.csv};
\addlegendentry{DeltaNet D=58 (104K) --- 60.1\%}

\addplot[color=transformer-pm, thick] table[x=epoch, y=test/acc, col sep=comma]
    {../report_data/transformer_L10_D29_H200_B64_warmup20_paramatch.csv};
\addlegendentry{Transformer D=29 (103K) --- 59.8\%}

\addplot[color=gfr-k4, thick] table[x=epoch, y=test/acc, col sep=comma]
    {../report_data/gfr_K4_D100_perslot.csv};
\addlegendentry{GFR K=4 per-slot (104K) --- 56.0\%}

\end{axis}
\end{tikzpicture}
\caption{Test accuracy for parameter-matched models ($\approx$103K). \textbf{APAN with RMSNorm keys achieves 64.2\%}---beating all other models by 4+ points. APN (prediction-error) reaches 58.3\%, while DeltaNet peaks early and declines. Both APAN variants show minimal overfitting (drop $<$0.3 points).}
\label{fig:paramatch}
\end{figure}

With only 105K parameters, APAN (RMSNorm) achieves \textbf{64.22\%} test accuracy---the highest among all $\sim$103K-parameter models, and competitive with the Transformer at 1.2M parameters (65.2\%).
Per 1K parameters, APAN extracts 0.61\% accuracy, versus 0.56\% for the prediction-error APN, 0.58\% for DeltaNet, and only 0.053\% for the Transformer.

\begin{table}[h]
\centering
\caption{Peak test accuracy for all models, sorted by performance. \textbf{APAN RMSNorm (105K params)} achieves 64.2\%---outperforming models with $3$--$12\times$ more parameters and within 1 point of the full Transformer at 1.2M.}
\label{tab:summary_ranked}
\begin{tabular}{clccc}
\toprule
\textbf{Rank} & \textbf{Model} & \textbf{Params} & \textbf{Peak Acc (\%)} & \textbf{Family} \\
\midrule
1 & Transformer H=4 & 1,205,610 & 65.16 & Transformer \\
\rowcolor{apan-rmsnorm!15}
\textbf{2} & \textbf{APAN RMSNorm (no FFN)} & \textbf{104,630} & \textbf{64.22} & \textbf{APAN} \\
3 & DeltaNet+FFN H=4 & 1,199,754 & 63.56 & DeltaNet \\
4 & Transformer H=1 & 1,205,610 & 63.50 & Transformer \\
5 & DeltaNet+FFN H=1 & 1,196,634 & 62.80 & DeltaNet \\
6 & DeltaNet D=100 & 304,610 & 62.28 & DeltaNet \\
7 & DeltaNet D=58 & 103,598 & 60.06 & DeltaNet \\
8 & Transformer D=29 & 102,554 & 59.75 & Transformer \\
9 & GFR+FFN D=115 & 1,196,720 & 58.93 & GFR \\
10 & APN act=none & 103,630 & 58.34 & APN \\
11 & APAN+FFN RMSNorm & 906,630 & 57.89 & APAN \\
12 & GFR+FFN D=34 & 105,974 & 57.57 & GFR \\
13 & APN+FFN & 1,196,720 & 56.84 & APN \\
14 & APAN tanh (no FFN) & 103,630 & 56.42 & APAN \\
15 & GFR K=4 & 103,690 & 56.00 & GFR \\
16 & GFR K=1 & 103,630 & 53.57 & GFR \\
\bottomrule
\end{tabular}
\end{table}

\begin{figure}[h]
\centering
\begin{tikzpicture}
\begin{axis}[
    width=0.85\textwidth,
    height=0.55\textwidth,
    xlabel={Epoch},
    ylabel={Test Accuracy (\%)},
    xmin=0, xmax=200,
    ymin=35, ymax=66,
    grid=major,
    grid style={gray!30},
    legend style={at={(0.98,0.02)}, anchor=south east, font=\footnotesize, draw=none, fill=white, fill opacity=0.8},
    every axis plot/.append style={thick},
    title={\textbf{All APAN Variants: RMSNorm vs.\ Tanh $\times$ FFN vs.\ No-FFN}},
    title style={font=\normalsize},
]

\addplot[color=apan-rmsnorm, thick] table[x=epoch, y=test/acc, col sep=comma]
    {../report_data/apan_L10_D100_rmsnorm_noFFN.csv};
\addlegendentry{APAN RMSNorm, no FFN (105K) --- \textbf{64.2\%}}

\addplot[color=apan-rmsnorm, thick, dashed] table[x=epoch, y=test/acc, col sep=comma]
    {../report_data/apan_L10_D100_rmsnorm_FFN.csv};
\addlegendentry{APAN RMSNorm + FFN (907K) --- 57.9\%}

\addplot[color=apan-tanh, thick] table[x=epoch, y=test/acc, col sep=comma]
    {../report_data/apan_L10_D100_tanh_noFFN.csv};
\addlegendentry{APAN tanh, no FFN (104K) --- 56.4\%}

\addplot[color=apan-tanh, thick, dashed] table[x=epoch, y=test/acc, col sep=comma]
    {../report_data/apan_L10_D100_tanh_FFN.csv};
\addlegendentry{APAN tanh + FFN (906K) --- 52.1\%}

\end{axis}
\end{tikzpicture}
\caption{All four APAN variants. \textbf{Key observations:} (1)~RMSNorm keys dominate tanh keys (+7.8 points without FFN, +5.8 with FFN). (2)~FFN \emph{always hurts} APAN ($-6.3$ points for RMSNorm, $-4.3$ for tanh). (3)~The best APAN (RMSNorm, no FFN) at 105K params outperforms the worst APAN+FFN variant (tanh+FFN at 906K) by 12.1 points. The FFN enables 100\% train accuracy but causes catastrophic overfitting.}
\label{fig:apan_variants}
\end{figure}

The comparison between APAN variants is illuminating:
\begin{itemize}
\item \textbf{APAN RMSNorm (64.22\%)}: The best. RMSNorm ensures $\|\mathbf{k}\|^2 = d$ exactly, giving the Hebbian update a consistent and well-scaled learning signal. The model does not reach 100\% train accuracy (76.6\%), indicating it is still learning rather than memorizing.
\item \textbf{APAN tanh (56.42\%)}: Significantly worse ($-7.8$ points). The tanh activation saturates gradients and compresses the key space, limiting the memory's discriminative power.
\item \textbf{APN act=none (58.34\%)}: The prediction-error variant with identity activation. Stable without normalization, but the weaker learning signal limits accuracy.
\end{itemize}

\textbf{Why RMSNorm keys help APAN so much.}
RMSNorm provides three benefits simultaneously: (1)~it guarantees stability ($\alpha < 1$ for any learned $\eta$), (2)~it removes the need for tanh's gradient-crushing activation, and (3)~it scales all keys to unit RMS, giving each time-step equal ``voice'' in the Hebbian update regardless of input magnitude.
The result is a Hebbian memory that learns consistently across all sequence positions.

% ============================================================================
\subsection{Key Finding 2: FFN Hurts APAN and APN but Helps DeltaNet and GFR}

\begin{figure}[h]
\centering
\begin{tikzpicture}
\begin{axis}[
    width=0.85\textwidth,
    height=0.55\textwidth,
    xlabel={Epoch},
    ylabel={Test Accuracy (\%)},
    xmin=0, xmax=200,
    ymin=35, ymax=67,
    grid=major,
    grid style={gray!30},
    legend style={at={(0.98,0.02)}, anchor=south east, font=\footnotesize, draw=none, fill=white, fill opacity=0.8},
    every axis plot/.append style={thick},
    title={\textbf{FFN-augmented group ($\approx$1.2M params)}},
    title style={font=\normalsize},
]

\addplot[color=transformer, thick] table[x=epoch, y=test/acc, col sep=comma]
    {../report_data/transformer_L10_D100_H200_B64_warmup20.csv};
\addlegendentry{Transformer D=100 (1.2M)}

\addplot[color=deltanet-ffn, thick] table[x=epoch, y=test/acc, col sep=comma]
    {../report_data/deltanet_L10_D104_H200_B64_ffn_warmup20_paramatch.csv};
\addlegendentry{DeltaNet+FFN D=104 (1.2M)}

\addplot[color=gfr-ffn, thick] table[x=epoch, y=test/acc, col sep=comma]
    {../report_data/gfr_K1_D115_ffn.csv};
\addlegendentry{GFR K=1+FFN D=115 (1.2M)}

\addplot[color=apn-ffn, thick] table[x=epoch, y=test/acc, col sep=comma]
    {../report_data/apn_L10_D115_H200_B64_ffn_warmup20_paramatch.csv};
\addlegendentry{APN+FFN D=115 (1.2M)}

\addplot[color=apan-rmsnorm, thick, dashed] table[x=epoch, y=test/acc, col sep=comma]
    {../report_data/apan_L10_D100_rmsnorm_FFN.csv};
\addlegendentry{APAN+FFN RMSNorm (907K)}

\addplot[color=apan-tanh, thick, dashed] table[x=epoch, y=test/acc, col sep=comma]
    {../report_data/apan_L10_D100_tanh_FFN.csv};
\addlegendentry{APAN+FFN tanh (906K)}

\addplot[color=apan-rmsnorm, thick, dotted] table[x=epoch, y=test/acc, col sep=comma]
    {../report_data/apan_L10_D100_rmsnorm_noFFN.csv};
\addlegendentry{APAN no-FFN RMSNorm (105K, reference)}

\end{axis}
\end{tikzpicture}
\caption{FFN-augmented models. DeltaNet+FFN (62.8\%) nearly matches the Transformer (63.5\%). GFR+FFN (58.9\%) benefits. Both APAN+FFN variants (dashed purple, 57.9\% and 52.1\%) \emph{underperform} plain APAN without FFN (dotted purple, 64.2\%)---the FFN enables memorization (100\% train acc) but destroys generalization. APN+FFN (56.8\%) also loses to plain APN (58.3\%).}
\label{fig:ffn}
\end{figure}

At $\approx$1.2M parameters, all FFN-augmented models reach 100\% training accuracy.
The critical difference is generalization:
\begin{itemize}
\item \textbf{DeltaNet+FFN (62.80\%)} nearly matches the Transformer (63.50\%), only 0.7 points behind.
\item \textbf{GFR+FFN (58.93\%)} benefits from the FFN, outperforming plain GFR $K{=}1$ by 5.4 points.
\item \textbf{APAN+FFN RMSNorm (57.89\%)}: Reaches 100\% train accuracy but badly overfits. \emph{Worse than APAN without FFN} (64.2\%) by 6.3 points despite $9\times$ more parameters.
\item \textbf{APN+FFN (56.84\%)}: Also worse than plain APN (58.34\%) by 1.5 points.
\item \textbf{APAN+FFN tanh (52.12\%)}: The worst FFN model---tanh + FFN combines the worst of both (compressed keys + overfitting). Final test loss reaches 7.8, indicating extreme overconfidence.
\end{itemize}
The FFN enables 100\% train accuracy for APAN/APN, but this capacity is entirely wasted on memorization.
The tied key-value design ($\mathbf{k} = \mathbf{v} = W\mathbf{x}$) produces representations too constrained for the FFN to refine productively.
In contrast, GFR's simpler element-wise recurrence produces features that the FFN can meaningfully transform.

\textbf{Lesson:} The APAN/APN architecture is best deployed \emph{without} FFN---its implicit regularization (not reaching 100\% train accuracy) is a feature, not a bug.

% ============================================================================
\subsection{Key Finding 3: APAN's Tanh $\eta$ Solves the Multi-Head Stability Problem}
\label{sec:nan}

\begin{figure}[h]
\centering
\begin{tikzpicture}
\begin{axis}[
    width=0.85\textwidth,
    height=0.55\textwidth,
    xlabel={Epoch},
    ylabel={Test Accuracy (\%)},
    xmin=0, xmax=200,
    ymin=35, ymax=67,
    grid=major,
    grid style={gray!30},
    legend style={at={(0.02,0.98)}, anchor=north west, font=\footnotesize, draw=none, fill=white, fill opacity=0.8},
    every axis plot/.append style={thick},
    title={\textbf{Multi-head ($H{=}4$) vs.\ single-head ($H{=}1$)}},
    title style={font=\normalsize},
]

\addplot[color=transformer, thick] table[x=epoch, y=test/acc, col sep=comma]
    {../report_data/transformer_D100_H4.csv};
\addlegendentry{Transformer H=4 (65.2\%)}

\addplot[color=transformer-pm, thick, dashed] table[x=epoch, y=test/acc, col sep=comma]
    {../report_data/transformer_L10_D100_H200_B64_warmup20.csv};
\addlegendentry{Transformer H=1 (63.5\%)}

\addplot[color=deltanet-ffn, thick] table[x=epoch, y=test/acc, col sep=comma]
    {../report_data/deltanet_D104_H4_ffn.csv};
\addlegendentry{DeltaNet+FFN H=4 (63.6\%)}

\addplot[color=deltanet-pm, thick, dashed] table[x=epoch, y=test/acc, col sep=comma]
    {../report_data/deltanet_L10_D104_H200_B64_ffn_warmup20_paramatch.csv};
\addlegendentry{DeltaNet+FFN H=1 (62.8\%)}

\addplot[color=apan-rmsnorm, thick] table[x=epoch, y=test/acc, col sep=comma]
    {../report_data/apan_L10_D100_H4_rmsnorm_noFFN.csv};
\addlegendentry{APAN H=4 (62.3\%)}

\addplot[color=apan-rmsnorm, thick, dashed] table[x=epoch, y=test/acc, col sep=comma]
    {../report_data/apan_L10_D100_rmsnorm_noFFN.csv};
\addlegendentry{APAN H=1 (64.2\%)}

\end{axis}
\end{tikzpicture}
\caption{Multi-head ($H{=}4$) vs.\ single-head. Transformer improves from 63.5\% to 65.2\% (+1.7 points). DeltaNet+FFN improves from 62.8\% to 63.6\% (+0.8 points). However, \textbf{APAN H=4 (62.3\%) is worse than APAN H=1 (64.2\%)} by 1.9 points---multi-head splits the $D{\times}D$ memory into $H$ smaller $25{\times}25$ matrices, reducing associative capacity. APN $H{=}4$ diverges to NaN (not shown).}
\label{fig:multihead}
\end{figure}

Multi-head attention ($H{=}4$) benefits the Transformer (+1.66 points, from 63.50\% to 65.16\%) and DeltaNet+FFN (+0.76 points, from 62.80\% to 63.56\%).
However, \textbf{multi-head hurts APAN}: APAN $H{=}4$ achieves 62.28\% peak (vs.\ 64.22\% for $H{=}1$), a \textbf{$-$1.94 point} drop.
Meanwhile, \textbf{both multi-head APN runs diverged completely}:

\begin{itemize}
\item \textbf{APAN $H{=}4$ RMSNorm} ($D{=}100$): Trains stably (tanh $\eta$ prevents divergence) but peaks at 62.28\% and overfits significantly (final: 58.09\%, drop = 4.19 points, train acc: 96.8\%). The 4 heads each operate on a $25{\times}25$ state matrix---less expressive than a single $100{\times}100$ matrix. Interestingly, several heads learn \emph{negative} $\eta$ (anti-Hebbian): layer~0 heads 0,1,2 have $\eta \approx -0.5, -0.6, -1.0$; layer~5 head~0 has $\eta = -0.98$. This suggests the model uses some heads for active forgetting.
\item \textbf{APN $H{=}4$ without FFN} ($D{=}100$): NaN from epoch 1. All scalars immediately become NaN.
\item \textbf{APN $H{=}4$ with FFN} ($D{=}116$): Trains for 9 epochs (peak 40.6\% at epoch 9), then NaN at epoch 10. The scalar logs reveal that head~3's $\eta$ (initialized at 0.1) \emph{goes negative} ($-0.12$ at epoch 9), causing the Hebbian update to amplify rather than correct errors. Meanwhile, head~0's $\eta$ (initialized at 1.0 with $\lambda{=}0.999$) grows unboundedly ($1.08$ at epoch 9).
\end{itemize}

\textbf{Root cause analysis.} With $H{=}4$, each head operates on a $D/H = 25$-dimensional subspace with a $25 \times 25$ state matrix.
The Hebbian outer-product update $\mathbf{k}_t \otimes (\mathbf{v}_t - S_{t-1}^\top \mathbf{k}_t)$ amplifies noise when:
(1)~the per-head dimension is too small, making the rank-1 update dominate the state;
(2)~$\eta$ is unconstrained and can go negative, reversing the sign of the delta correction;
(3)~$\lambda{=}0.9$ (head~2's initialization) provides insufficient damping, so errors accumulate rapidly.
For the no-FFN variant, the $25{\times}25$ state is so small that the system is unstable from the first gradient step.
These issues do not affect DeltaNet because its $\beta_t = \sigma(\cdot) \in (0,1)$ is always positive and bounded.

\textbf{APAN's solution.} The APAN architecture addresses the NaN problem with two mechanisms: (1)~$\eta = \tanh(\hat\eta) \in (-1, 1)$ is bounded by construction---it cannot diverge; (2)~RMSNorm key normalization ensures $\|\mathbf{k}\|^2 = d$ exactly, making the stability condition $|\alpha| = |\lambda + \eta(1-\lambda)| < 1$ hold unconditionally for any $\eta \in (-1,1)$ and $\lambda \in (0,1)$.
APAN $H{=}4$ confirms this: it trains stably for all 200 epochs with no NaN.
However, stability $\ne$ optimality: splitting the $100{\times}100$ state into four $25{\times}25$ matrices reduces the rank of each head's memory from 100 to 25, limiting its ability to store complex associations.
The result is a model that overfits more (96.8\% train, 4.2-point drop) and generalizes less (62.3\% vs.\ 64.2\%).

\textbf{Why multi-head hurts APAN but helps the Transformer.}
The Transformer benefits from multiple heads because each head attends to different parts of the sequence (different ``what to attend to'' patterns), and the FFN combines their outputs.
APAN's $D{\times}D$ matrix state stores associative memories; splitting it into $H$ smaller matrices reduces the capacity of each memory without a mechanism to combine them constructively.
At $D=100, H=4$, each head has a $25{\times}25$ matrix with rank at most 25---a $4\times$ reduction in representational capacity compared to the single-head $100{\times}100$ matrix.

% ============================================================================
\subsection{Key Finding 4: GFR Validates the Importance of APN's Matrix State}

\begin{figure}[h]
\centering
\begin{tikzpicture}
\begin{axis}[
    width=0.85\textwidth,
    height=0.55\textwidth,
    xlabel={Epoch},
    ylabel={Test Accuracy (\%)},
    xmin=0, xmax=200,
    ymin=35, ymax=60,
    grid=major,
    grid style={gray!30},
    legend style={at={(0.98,0.02)}, anchor=south east, font=\footnotesize, draw=none, fill=white, fill opacity=0.8},
    every axis plot/.append style={thick},
    title={\textbf{APN vs.\ GFR variants ($\approx$103K params, no FFN)}},
    title style={font=\normalsize},
]

\addplot[color=apn-none, thick] table[x=epoch, y=test/acc, col sep=comma]
    {../report_data/apn_L10_D100_H200_B64_act-none_warmup20.csv};
\addlegendentry{APN act=none (58.3\%)}

\addplot[color=gfr-k4, thick] table[x=epoch, y=test/acc, col sep=comma]
    {../report_data/gfr_K4_D100_perslot.csv};
\addlegendentry{GFR K=4 per-slot (56.0\%)}

\addplot[color=gfr-k4, thick, dashed] table[x=epoch, y=test/acc, col sep=comma]
    {../report_data/gfr_K4_D100_perneuron.csv};
\addlegendentry{GFR K=4 per-neuron (56.2\%)}

\addplot[color=gfr-k1, thick] table[x=epoch, y=test/acc, col sep=comma]
    {../report_data/gfr_K1_D100.csv};
\addlegendentry{GFR K=1 per-slot (53.6\%)}

\addplot[color=gfr-k1, thick, dashed] table[x=epoch, y=test/acc, col sep=comma]
    {../report_data/gfr_K1_D100_perneuron.csv};
\addlegendentry{GFR K=1 per-neuron (52.4\%)}

\end{axis}
\end{tikzpicture}
\caption{APN vs.\ GFR variants at matched parameters. APN's $D{\times}D$ matrix state outperforms GFR's scalar memory slots by 2.3--5.9 points. Adding slots ($K{=}4$) helps significantly (+2.4 points over $K{=}1$), partially compensating for the lack of associative capacity. Per-neuron parameterization offers no clear advantage over per-slot.}
\label{fig:gfr}
\end{figure}

GFR serves as an ablation of APN, replacing the $D \times D$ matrix state with $K$ independent exponentially-decaying scalar memories.
The results confirm that \textbf{APN's matrix state provides meaningful associative capacity}:
\begin{itemize}
\item GFR $K{=}1$: 53.57\% ($-4.8$ points vs.\ APN). A single exponential decay cannot perform the delta-rule error correction.
\item GFR $K{=}4$ per-slot: 56.00\% ($-2.3$ points). Multiple timescales partially compensate by allowing the model to store short- and long-term features simultaneously.
\item GFR $K{=}4$ per-neuron: 56.21\% ($-2.1$ points). Per-neuron parameters (800 extra per layer) do not meaningfully outperform the simpler per-slot parameterization (8 extra per layer).
\item GFR $K{=}1$ per-neuron: 52.43\% ($-5.9$ points). Worse than per-slot $K{=}1$, suggesting that per-neuron decay rates with a single slot overfit.
\end{itemize}

GFR is significantly faster than APN: 14.0 s/epoch for $K{=}1$ (vs.\ 25.5 s for APN) and 22.7 s for $K{=}4$.
The HGRN kernel's element-wise scan is computationally trivial compared to APN's tiled $D \times D$ delta-rule computation.

% ============================================================================
\subsection{Key Finding 5: Loss--Accuracy Divergence in Overparameterized Models}
\label{sec:loss_divergence}

\begin{figure}[h]
\centering
\begin{subfigure}[b]{0.48\textwidth}
\begin{tikzpicture}
\begin{axis}[
    width=\textwidth,
    height=0.75\textwidth,
    xlabel={Epoch},
    ylabel={Test Accuracy (\%)},
    xmin=0, xmax=200,
    ymin=35, ymax=67,
    grid=major,
    grid style={gray!30},
    legend style={at={(0.98,0.02)}, anchor=south east, font=\tiny, draw=none, fill=white, fill opacity=0.8},
    every axis plot/.append style={thick},
]
\addplot[color=transformer, thick] table[x=epoch, y=test/acc, col sep=comma]
    {../report_data/transformer_D100_H4.csv};
\addlegendentry{Transformer H=4}
\addplot[color=deltanet-ffn, thick] table[x=epoch, y=test/acc, col sep=comma]
    {../report_data/deltanet_D104_H4_ffn.csv};
\addlegendentry{DeltaNet+FFN H=4}
\addplot[color=gfr-ffn, thick] table[x=epoch, y=test/acc, col sep=comma]
    {../report_data/gfr_K1_D115_ffn.csv};
\addlegendentry{GFR+FFN}
\end{axis}
\end{tikzpicture}
\caption{Test accuracy continues improving}
\end{subfigure}
\hfill
\begin{subfigure}[b]{0.48\textwidth}
\begin{tikzpicture}
\begin{axis}[
    width=\textwidth,
    height=0.75\textwidth,
    xlabel={Epoch},
    ylabel={Test Loss (CE)},
    xmin=0, xmax=200,
    ymin=0.5, ymax=6.0,
    grid=major,
    grid style={gray!30},
    legend style={at={(0.02,0.98)}, anchor=north west, font=\tiny, draw=none, fill=white, fill opacity=0.8},
    every axis plot/.append style={thick},
]
\addplot[color=transformer, thick] table[x=epoch, y=test/loss, col sep=comma]
    {../report_data/transformer_D100_H4.csv};
\addlegendentry{Transformer H=4}
\addplot[color=deltanet-ffn, thick] table[x=epoch, y=test/loss, col sep=comma]
    {../report_data/deltanet_D104_H4_ffn.csv};
\addlegendentry{DeltaNet+FFN H=4}
\addplot[color=gfr-ffn, thick] table[x=epoch, y=test/loss, col sep=comma]
    {../report_data/gfr_K1_D115_ffn.csv};
\addlegendentry{GFR+FFN}
\end{axis}
\end{tikzpicture}
\caption{Test loss diverges to $\sim$5.0--5.5}
\end{subfigure}
\caption{Loss--accuracy divergence in overparameterized models. All three models reach 100\% train accuracy and develop increasingly confident predictions. On test samples where the model is wrong, the large logit magnitudes produce enormous cross-entropy loss. Since accuracy only measures correct/incorrect (not confidence), test accuracy can improve even as test loss explodes. Final test losses: Transformer H=4: 5.49, DeltaNet+FFN H=4: 5.19, GFR+FFN: 5.31.}
\label{fig:loss_divergence}
\end{figure}

A striking phenomenon in our experiments: models that achieve 100\% training accuracy show test loss climbing to $\sim$5.0--5.5 in later epochs, while test accuracy remains stable or even improves slightly.
This occurs in the Transformer ($H{=}4$: final test loss 5.49, acc 65.1\%), DeltaNet+FFN ($H{=}4$: loss 5.19, acc 63.4\%), and GFR+FFN (loss 5.31, acc 58.8\%).

\textbf{Explanation.} Cross-entropy loss is $-\log p_{\text{correct}}$.
When a model reaches 100\% train accuracy with near-zero train loss ($\sim 10^{-9}$), its logits become very large in magnitude.
On test samples, the model is still correct on $\sim$63--65\% of examples (producing near-zero per-sample loss for those), but for the $\sim$35\% of misclassified samples, the large logit magnitudes mean the model is \emph{confidently wrong}, yielding per-sample losses of $\sim$15--20.
The mean across all samples produces the observed $\sim$5.0 test loss.
Accuracy only counts the argmax and is insensitive to confidence, so it remains stable.

\textbf{Notably, APN and GFR without FFN never exhibit this divergence} (final test losses $\sim$1.2--1.3) because they cannot reach 100\% train accuracy---their implicit regularization prevents the logit magnitudes from growing unboundedly.

% ============================================================================
\subsection{Speed and Memory Comparison}

\begin{figure}[h]
\centering
\begin{tikzpicture}
\begin{axis}[
    ybar,
    width=0.95\textwidth,
    height=0.4\textwidth,
    ylabel={Epoch Time (seconds)},
    symbolic x coords={GFR K1\\D100, GFR K4\\D100, APN\\D100, Transf.\\D29, Transf.\\D100, GFR+FFN\\D115, APN+FFN\\D115, DN+FFN\\D104, DN\\D58, Transf.H4\\D100, DN\\D100},
    xtick=data,
    x tick label style={align=center, font=\scriptsize},
    ymin=0, ymax=120,
    bar width=12pt,
    nodes near coords,
    nodes near coords style={font=\tiny, above},
    grid=major,
    grid style={gray!30, dashed},
]
\addplot[fill=gfr-k1!80] coordinates {
    (GFR K1\\D100, 14.0)
    (GFR K4\\D100, 22.7)
    (APN\\D100, 25.5)
    (Transf.\\D29, 32.6)
    (Transf.\\D100, 39.3)
    (GFR+FFN\\D115, 33.3)
    (APN+FFN\\D115, 55.3)
    (DN+FFN\\D104, 112.1)
    (DN\\D58, 69.0)
    (Transf.H4\\D100, 85.1)
    (DN\\D100, 100.4)
};
\end{axis}
\end{tikzpicture}
\caption{Training time per epoch (lower is better). GFR $K{=}1$ is the fastest (14s), followed by GFR $K{=}4$ (23s) and APN (26s). DeltaNet is consistently the slowest due to its Triton kernel overhead at these dimensions.}
\label{fig:epoch_time}
\end{figure}

\begin{table}[h]
\centering
\caption{Speed and memory comparison for key configurations.}
\label{tab:speed_memory}
\begin{tabular}{lccc}
\toprule
\textbf{Model (params)} & \textbf{s/epoch} & \textbf{Peak GPU (MB)} & \textbf{Relative speed} \\
\midrule
GFR $K{=}1$ $D{=}100$ (104K) & 14.0 & 1,460 & 1.0$\times$ (fastest) \\
GFR $K{=}4$ $D{=}100$ (104K) & 22.7 & 3,154 & 1.6$\times$ \\
APN $D{=}100$ (104K) & 25.5 & 1,590 & 1.8$\times$ \\
Transformer $D{=}29$ (103K) & 32.6 & 4,623 & 2.3$\times$ \\
Transformer $D{=}100$ (1.2M) & 39.3 & 4,278 & 2.8$\times$ \\
DeltaNet $D{=}58$ (104K) & 69.0 & 1,151 & 4.9$\times$ \\
DeltaNet $D{=}100$ (305K) & 100.4 & 2,133 & 7.2$\times$ \\
Transformer $H{=}4$ $D{=}100$ (1.2M) & 85.1 & 17,158 & 6.1$\times$ \\
\bottomrule
\end{tabular}
\end{table}

Key observations:
\begin{itemize}
\item \textbf{GFR $K{=}1$ is fastest} at 14s/epoch---$1.8\times$ faster than APN and $7\times$ faster than DeltaNet.
\item \textbf{APN is 4$\times$ faster than DeltaNet} at the same width ($D{=}100$), due to its single projection vs.\ three.
\item \textbf{Transformer $H{=}4$ uses 17 GB} GPU memory---$4\times$ more than $H{=}1$ (4.3 GB)---because FlashAttention-2's $H{=}4$ configuration stores more intermediate activations for the backward pass.
\item \textbf{GFR $K{=}4$ uses 3.2 GB}---$2\times$ more than $K{=}1$ (1.5 GB)---proportional to the $4\times$ larger state vector.
\end{itemize}

% ============================================================================
\subsection{APN Scalar Dynamics}

\begin{figure}[h]
\centering
\begin{subfigure}[b]{0.48\textwidth}
\begin{tikzpicture}
\begin{axis}[
    width=\textwidth,
    height=0.7\textwidth,
    xlabel={Epoch},
    ylabel={$\eta$},
    xmin=0, xmax=200,
    grid=major,
    grid style={gray!30},
    legend style={at={(0.02,0.98)}, anchor=north west, font=\footnotesize, draw=none, fill=white, fill opacity=0.8},
    every axis plot/.append style={thick},
]
\addplot[color=apn-none, thick] table[x=epoch, y=apn/eta_layer0, col sep=comma]
    {../report_data/apn_L10_D100_H200_B64_act-none_warmup20_scalars.csv};
\addlegendentry{Layer 0}
\addplot[color=transformer, thick, dashed] table[x=epoch, y=apn/eta_layer9, col sep=comma]
    {../report_data/apn_L10_D100_H200_B64_act-none_warmup20_scalars.csv};
\addlegendentry{Layer 9}
\end{axis}
\end{tikzpicture}
\caption{Learned $\eta$ (APN act=none)}
\end{subfigure}
\hfill
\begin{subfigure}[b]{0.48\textwidth}
\begin{tikzpicture}
\begin{axis}[
    width=\textwidth,
    height=0.7\textwidth,
    xlabel={Epoch},
    ylabel={$\lambda$},
    xmin=0, xmax=200,
    ymin=0.990, ymax=1.001,
    grid=major,
    grid style={gray!30},
    legend style={at={(0.02,0.02)}, anchor=south west, font=\footnotesize, draw=none, fill=white, fill opacity=0.8},
    every axis plot/.append style={thick},
]
\addplot[color=apn-none, thick] table[x=epoch, y=apn/lam_layer0, col sep=comma]
    {../report_data/apn_L10_D100_H200_B64_act-none_warmup20_scalars.csv};
\addlegendentry{Layer 0}
\addplot[color=transformer, thick, dashed] table[x=epoch, y=apn/lam_layer9, col sep=comma]
    {../report_data/apn_L10_D100_H200_B64_act-none_warmup20_scalars.csv};
\addlegendentry{Layer 9}
\end{axis}
\end{tikzpicture}
\caption{Learned $\lambda$ (APN act=none)}
\end{subfigure}
\caption{Evolution of APN scalars during training. Both $\eta$ and $\lambda$ remain close to their initializations ($\eta{=}1.0$, $\lambda{=}0.999$), confirming the initialization is near-optimal. $\lambda$ converges toward 1.0 (minimal forgetting), consistent with the Seq-CIFAR-10 task requiring information from the entire 1024-step sequence.}
\label{fig:scalars}
\end{figure}

% ============================================================================
\section{Activation Function Analysis}

\begin{table}[h]
\centering
\caption{Effect of activation function on APN performance ($D=100$, $L=10$, 200 epochs).}
\label{tab:activations}
\begin{tabular}{lcccc}
\toprule
\textbf{Activation} & \textbf{Best Test Acc} & \textbf{Best Ep} & \textbf{Train Acc (final)} & \textbf{Notes} \\
\midrule
\texttt{none} (identity, warmup) & \textbf{58.34\%} & 189 & 62.92\% & Best; no gradient saturation \\
\texttt{none} (no warmup) & 58.12\% & 163 & 61.84\% & Slightly worse without warmup \\
\texttt{tanh} (better init) & 53.15\% & 168 & 70.24\% & Overfits more; bounded keys \\
\texttt{tanh} (default init) & 51.69\% & 162 & 62.05\% & Baseline \\
\texttt{tanh} (frozen $\lambda$) & 49.54\% & --- & 61.21\% & Freezing $\lambda$ hurts \\
\bottomrule
\end{tabular}
\end{table}

The identity activation (\texttt{none}) outperforms \texttt{tanh} by 5--6 points.
\texttt{tanh} saturates for large inputs, compressing the key space and limiting the memory matrix's ability to distinguish similar inputs.
Notably, the best \texttt{tanh} variant (better init, 53.15\%) overfits more heavily: 70.24\% final train accuracy vs.\ only 62.92\% for \texttt{act=none}.

% ============================================================================
\section{Computational Complexity}

\begin{table}[h]
\centering
\caption{Theoretical complexity per layer. $T$: sequence length, $D$: hidden dim, $C$: chunk size (64).}
\label{tab:complexity}
\begin{tabular}{lcc}
\toprule
\textbf{Model} & \textbf{Time} & \textbf{Memory (state)} \\
\midrule
Transformer (softmax) & $O(T^2 D + T D^2)$ & $O(T^2)$ per head${}^\dagger$ \\
DeltaNet / APN (chunked) & $O(TCD + TD^2/C)$ & $O(D^2)$ per head \\
DeltaNet / APN (recurrent) & $O(TD^2)$ & $O(D^2)$ per head \\
GFR (HGRN, chunked) & $O(TDK)$ & $O(DK)$ \\
\bottomrule
\end{tabular}

\smallskip
{\footnotesize ${}^\dagger$ FlashAttention-2 reduces HBM traffic to $O(T)$ but FLOPs remain $O(T^2)$.}
\end{table}

\textbf{Why APN/DeltaNet can be faster.}
For $T=1024$, $D=100$: the softmax attention FLOPs scale as $T^2 D = 1024^2 \times 100 \approx 10^8$, while the chunked delta-rule scales as $TCD = 1024 \times 64 \times 100 \approx 6.5 \times 10^6$ --- a $\approx 15\times$ reduction in the attention-like computation.
However, the APN/DeltaNet kernels are Triton-based (less mature than FlashAttention's hand-tuned CUDA), and at small $D$ the linear projections dominate, so the wall-clock advantage depends on the specific configuration.

For \textbf{inference}, the advantage is even clearer: APN/DeltaNet/GFR maintain a fixed-size state (constant memory), while the Transformer's KV cache grows linearly with sequence length.

% ============================================================================
\section{GPU Kernel Implementation and the Chunk Method}
\label{sec:kernels}

A critical contributor to APN/DeltaNet/GFR's practical speed is the \textbf{fused Triton kernel} implementation from the Flash Linear Attention (FLA) library.
This section explains the GPU execution model, contrasts a naive PyTorch implementation with the kernel-based approach, and details the chunk algorithm.

\subsection{GPU Memory Hierarchy}

Modern GPUs have a layered memory system, where each level trades capacity for speed:

\begin{table}[h]
\centering
\caption{GPU memory hierarchy (H200).}
\label{tab:memhierarchy}
\begin{tabular}{lccc}
\toprule
\textbf{Level} & \textbf{Size} & \textbf{Bandwidth} & \textbf{Control} \\
\midrule
Registers & $\approx$256 KB / SM & Instantaneous & Per-thread \\
Shared Memory (SRAM) & 228 KB / SM & $\approx$20 TB/s & Programmer-managed \\
L2 Cache & 40--60 MB & $\approx$6 TB/s & Automatic \\
HBM (Global Memory) & 141 GB & $\approx$3 TB/s & Default storage \\
\bottomrule
\end{tabular}
\end{table}

Most GPU workloads are \textbf{memory-bound}: the arithmetic units can perform trillions of operations per second, but spend most time waiting for data from HBM.
The key to fast kernels is keeping data in shared memory / registers as long as possible to minimize HBM round-trips.

\subsection{Naive APN vs.\ Fused Kernel APN}

\textbf{Naive PyTorch implementation.} A straightforward APN loop issues 7+ kernel launches per timestep (activation, two matmuls, scale, outer product, add, copy):
\begin{itemize}
\item $T \times 7 = 7{,}168$ kernel launches per layer, each with $\sim$10\,$\mu$s overhead $\approx$ 72\,ms in launch overhead alone.
\item The $D \times D$ state matrix $S$ is read from / written to HBM at every timestep.
\item A Python for-loop serializes all timesteps, preventing the GPU from planning work ahead.
\item Backpropagation requires saving all $T$ intermediate states. For $B=64$, $D=100$, $T=1024$, $L=10$:
\[
B \times T \times D^2 \times 4\,\text{bytes} \times L = 64 \times 1024 \times 10^4 \times 4 \times 10 \approx 26\,\text{GB}.
\]
\end{itemize}

\textbf{Fused kernel implementation.}
The FLA library replaces this loop with compact Triton kernel pipelines:
\begin{itemize}
\item \textbf{APN}: \texttt{chunk\_gated\_delta\_rule} --- a multi-kernel pipeline computing gates, intra-chunk WY terms, inter-chunk state propagation, and outputs.
\item \textbf{GFR}: \texttt{chunk\_hgrn} --- simple element-wise gated prefix scan (trivially parallel within chunks).
\item \textbf{DeltaNet}: \texttt{fused\_recurrent\_delta\_rule} --- a single-kernel sequential scan.
\end{itemize}

The improvements are:
\begin{enumerate}
\item \textbf{A handful of kernel launches} instead of 7{,}168, eliminating launch overhead.
\item \textbf{State tiles stay on-chip}: each Triton program owns one batch/head/tile of the $D \times D$ state in registers or SRAM. Between timesteps, the state is updated in-place without an HBM round-trip.
\item \textbf{No Python loop}: the entire forward pass executes autonomously on the GPU.
\item \textbf{Smart backward pass}: the chunk method avoids storing all $T$ intermediate states (see below).
\end{enumerate}
This is why the kernel-based APN is $10$--$100\times$ faster than the naive version and uses $\sim$1.6\,GB instead of $\sim$26\,GB.

\subsection{The Chunk Algorithm}
\label{sec:chunk}

The delta-rule recurrence $S_t = f(S_{t-1}, \mathbf{k}_t, \mathbf{v}_t)$ is inherently sequential: $S_t$ depends on $S_{t-1}$.
A fused recurrent kernel loops through all $T$ timesteps one by one; parallelism is limited to batch and state dimensions.

The \textbf{chunk method} splits the sequence into $T/C$ chunks of $C$ timesteps (default $C=64$) and restructures the computation:

\begin{enumerate}
\item \textbf{Intra-chunk (parallel across chunks).}
Within each chunk, the recurrence is reformulated as a lower-triangular linear system using the \textbf{WY representation}:
\[
L \, \tilde{\mathbf{v}} = \mathbf{v} - S_{\text{prev}} K^\top, \quad
L_{ij} = \begin{cases} 1 & i = j \\ -\beta_i \, \mathbf{k}_i^\top \mathbf{k}_j \prod_{m=j+1}^{i-1} g_m & i > j \\ 0 & i < j \end{cases}
\]
This triangular solve maps to GPU matrix operations and can be solved for all $T/C$ chunks simultaneously.

\item \textbf{Inter-chunk (sequential across chunk boundaries).}
The $D \times D$ state is propagated from chunk 0 $\to$ chunk 1 $\to \cdots \to$ chunk $T/C-1$.
For $T=1024, C=64$, this is only 16 sequential state updates instead of 1024.
\end{enumerate}

\begin{table}[h]
\centering
\caption{Comparison of execution strategies for $T=1024$, $C=64$.}
\label{tab:strategies}
\begin{tabular}{lccc}
\toprule
\textbf{Method} & \textbf{Sequential depth} & \textbf{HBM reads of $S$} & \textbf{Backward state storage} \\
\midrule
Naive Python loop & $T = 1024$ & $T \times D^2$ & All $T$ states \\
Fused recurrent & $T = 1024$ & 1 (load once) & Recompute \\
Chunk ($C=64$) & $T/C = 16$ & $T/C \times D^2$ & $T/C$ boundary states \\
\bottomrule
\end{tabular}
\end{table}

The chunk method's advantage grows with sequence length: at $T=8192$, the sequential depth is 128 (vs.\ 8192); at $T=65536$ it is 1024 (vs.\ 65536).
For the backward pass, only $T/C$ boundary states need to be stored---a $64\times$ reduction compared to the naive approach.

\textbf{GFR's simplicity.} The HGRN kernel used by GFR performs an element-wise gated scan (each neuron independently), which is trivially parallelizable within chunks and requires no WY transform or triangular solve. This explains why GFR $K{=}1$ (14\,s/epoch) is even faster than APN (25.5\,s/epoch) despite using the same chunk infrastructure.

\subsection{Tile Alignment and Performance Cliffs}

The chunk kernel processes the $D \times D$ state in tiles (typically $64 \times 64$ blocks).
When $D$ is not a multiple of the tile size, the kernel must pad to the next tile boundary, wasting compute:

\begin{table}[h]
\centering
\caption{Effect of $D$ on tile count and efficiency (tile size $=64$).}
\label{tab:tiling}
\begin{tabular}{ccccc}
\toprule
$D$ & Tiles/dim & Tile blocks & Padding waste & Relative efficiency \\
\midrule
64 & 1 & 1 & 0\% & Best \\
100 & 2 & 4 & 22\% & Good \\
128 & 2 & 4 & 0\% & Best \\
173 & 3 & 9 & 10\% & Poor (tile jump) \\
256 & 4 & 16 & 0\% & Good \\
\bottomrule
\end{tabular}
\end{table}

For $D=100$: $\lceil 100/64 \rceil = 2$ tiles per dimension $\to 4$ tile blocks.
For $D=173$: $\lceil 173/64 \rceil = 3$ tiles $\to 9$ blocks, plus padding waste and reduced occupancy.
This explains the $\approx$18$\times$ slowdown from $D=100$ (25.5\,s/epoch) to $D=173$ (452.8\,s/epoch) observed in our APN experiments---far worse than the $1.73^2 \approx 3\times$ ratio of $D^2$ arithmetic alone.

\textbf{Recommendation:} For efficient Triton kernel execution, prefer tile-aligned dimensions $D \in \{64, 128, 192, 256\}$ or small dimensions with only 2 tiles per axis ($D \le 128$).

\subsection{Why APN Is Faster Than DeltaNet and the Transformer}

Beyond the shared kernel infrastructure, three factors explain APN's speed advantage:

\begin{enumerate}
\item \textbf{Fewer projections.}
APN uses a single $D \times D$ projection per layer; DeltaNet uses three ($W_Q, W_K, W_V$); the Transformer uses four projections plus a $D \to 4D \to D$ FFN.
At $D=100$: APN computes 10K multiply-adds per token per layer; DeltaNet 30K; the Transformer 120K.

\item \textbf{Compact recurrent state.}
APN/DeltaNet store a $D \times D$ state ($\approx$40\,KB at $D=100$ in FP32).
The Transformer must store Q/K/V activations proportional to $T$ for the backward pass, even with FlashAttention-2's memory-efficient recomputation, resulting in higher HBM traffic.

\item \textbf{Linear vs.\ quadratic complexity in $T$.}
Softmax attention (even with FlashAttention) is $O(T^2 D)$; the chunked delta rule is $O(TCD + TD^2/C)$---linear in $T$.
At $T=1024, D=100$: attention FLOPs are $\approx 15\times$ higher for the Transformer.
\end{enumerate}

\textbf{Caveat:} FlashAttention-2 is implemented in hand-tuned CUDA, achieving near-peak throughput.
The FLA kernels are Triton-based (correct but not fully optimized for Hopper).
As $T$ grows beyond 4096, the $O(T)$ vs.\ $O(T^2)$ gap dominates regardless of kernel maturity.

\textbf{Note on Triton version:} We pin \texttt{triton>=3.3.0,<3.4.0} due to a known correctness bug in the gated chunk backward kernel on Hopper GPUs with Triton $\ge$3.4 (FLA issue \#640).
Triton 3.3 generates correct code and runs \emph{faster} on H200 (25.5\,s/epoch for APN) than on A100 with Triton 3.6, as Hopper's raw hardware advantages (more SMs, higher bandwidth, 228\,KB shared memory) compensate for missing compiler optimizations.

% ============================================================================
\section{Discussion}

\subsection{Why APAN Outperforms APN: Attractor vs.\ Prediction-Error Dynamics}

The central finding of this report is that APAN's Hebbian attractor update ($\mathbf{v} + S^\top\mathbf{k}$) outperforms APN's prediction-error update ($\mathbf{v} - S^\top\mathbf{k}$) by 5.9 points at matched parameters.
The explanation lies in the dynamics:

\begin{itemize}
\item \textbf{Prediction-error (APN)}: The update $\mathbf{k} \otimes (\mathbf{v} - S^\top\mathbf{k})$ self-extinguishes as $S$ converges to encode the association $S^\top\mathbf{k} \approx \mathbf{v}$. This is stable but conservative---the memory stores each pattern weakly and equally.

\item \textbf{Attractor (APAN)}: The update $\mathbf{k} \otimes (\mathbf{v} + S^\top\mathbf{k})$ reinforces existing memories. Patterns that are partially recalled are strengthened further, creating deeper basins of attraction. This allows the model to build up strong, persistent memories over the 1024-step sequence.
\end{itemize}

The key risk of the attractor update---instability---is fully mitigated by RMSNorm keys + tanh $\eta$, which guarantee $\alpha \in (0.98, 1.0)$ for $\lambda = 0.99$.
The result is a system that combines the representational power of positive feedback with the safety of bounded gains.

\subsection{The Critical Role of Key Normalization}

APAN+RMSNorm (64.2\%) vs.\ APAN+tanh (56.4\%) shows that key normalization is the single most important design choice.
The tanh activation has two problems:
(1)~it saturates for large inputs, compressing the key space and limiting discrimination between similar inputs;
(2)~its gradient vanishes at saturation, preventing the model from learning to adjust the keys.
RMSNorm avoids both issues: it normalizes keys to unit RMS without crushing gradients, and it ensures consistent $\|\mathbf{k}\|^2 = d$ so that the learning rate $\eta$ has a predictable effect regardless of input magnitude.

\subsection{Why FFN Hurts Hebbian Models}

Both APAN and APN perform worse with FFN augmentation.
The fundamental issue is that the Hebbian/delta-rule update stores associations of the form $\mathbf{k}_t \otimes \mathbf{h}_t$---the memory operates in the \emph{same feature space} as the input.
An FFN applied after the layer transforms the features into a different space, but the next layer's Hebbian memory still operates on the original key space.
This creates a ``representation mismatch'' where the FFN's transformations are not coherent with the memory's storage format.
In contrast, GFR and DeltaNet (with separate Q/K/V projections) can adapt their representations to work with the FFN.

\subsection{GFR as a Minimal Baseline}

GFR demonstrates that even a trivial element-wise gated recurrence (14 s/epoch) can achieve 53--56\% accuracy---only 2--8 points below APAN.
The gap confirms that APAN's $D{\times}D$ matrix state provides genuine associative capacity (the ability to store and retrieve key-value associations via the Hebbian rule), not just temporal integration.

% ============================================================================
\section{Conclusions}

\begin{enumerate}
\item \textbf{APAN with RMSNorm is the best model at $\sim$103K parameters} (64.2\%)---outperforming all other architectures including prediction-error APN (58.3\%), DeltaNet (60.1\%), and the Transformer (59.8\%) at matched parameters.
It is competitive with the Transformer at 1.2M parameters (65.2\%) using $12\times$ fewer parameters.

\item \textbf{The Hebbian attractor update outperforms the prediction-error (delta) rule.}
APAN's learning signal ($\mathbf{v} + S^\top\mathbf{k}$, the full output) is stronger than APN's ($\mathbf{v} - S^\top\mathbf{k}$, the prediction error), yielding +5.9 points at matched parameters.
The attractor dynamics create deeper memory traces that persist across the full 1024-step sequence.

\item \textbf{RMSNorm key normalization is critical for APAN.}
APAN+RMSNorm (64.2\%) outperforms APAN+tanh (56.4\%) by 7.8 points.
RMSNorm ensures stability without compressing gradients, and gives each time-step equal weight in the update.

\item \textbf{APAN is implemented on the unmodified delta-rule kernel} by negating $v$ and $\beta$---a one-line change in the calling code, with zero modifications to the optimized Triton kernel.

\item \textbf{FFN augmentation hurts APAN and APN.}
APAN+FFN (57.9\%) is 6.3 points worse than plain APAN (64.2\%); APN+FFN (56.8\%) is 1.5 points worse than plain APN (58.3\%).
The tied key-value design already provides sufficient representational capacity; FFN only enables overfitting.

\item \textbf{APN (prediction-error) is uniquely resistant to overfitting} (0.17-point peak-to-final drop), requiring no early stopping or regularization.

\item \textbf{Multi-head APN ($H{=}4$) diverges} due to unconstrained $\eta$ going negative. APAN's tanh-bounded $\eta$ and RMSNorm keys prevent this failure mode.

\item \textbf{GFR validates the importance of the matrix state}: the $D{\times}D$ associative memory provides 2--8 points over scalar exponential decay at equal parameters.

\item \textbf{DeltaNet+FFN $H{=}4$ (63.6\%) nearly matches Transformer $H{=}4$ (65.2\%)} with linear-time complexity---the practical gap is only 1.6 points.

\item \textbf{The identity activation (\texttt{act=none}) combined with RMSNorm is the best key processing strategy}, outperforming tanh by 7.8 points (APAN) or 5--6 points (APN).
\end{enumerate}

\subsection{Future Directions}

\begin{enumerate}
\item \textbf{Multi-head APAN}: With tanh $\eta$ and RMSNorm keys, multi-head APAN should be stable. Test $H{=}4$ at $D{=}100$.
\item \textbf{Separate key/value projections}: Move from $D^2$ to $2D^2$ parameters to decouple the representation bottleneck.
\item \textbf{Longer sequences}: $T > 4096$ where linear-time complexity becomes decisive and the Hebbian memory's associative capacity should show greater advantages.
\item \textbf{Language modeling}: Evaluate on autoregressive tasks where associative memory may show greater advantages.
\item \textbf{Negative $\eta$ analysis}: APAN's tanh parameterization allows anti-Hebbian learning ($\eta < 0$). Investigate whether the model learns to use this.
\end{enumerate}

% ============================================================================
\clearpage
\appendix
\section{Complete Training Curves for All Runs}
\label{app:all_curves}

This appendix presents test accuracy and test loss curves for all 24 successfully-trained runs (excluding the two NaN-diverged APN $H{=}4$ configurations).

\subsection{Group 1: $\approx$1.2M Parameter Models}

All models in this group reach 100\% training accuracy. The key differentiator is test generalization and the loss--accuracy divergence pattern (Section~\ref{sec:loss_divergence}).

\begin{figure}[h]
\centering
\begin{subfigure}[b]{0.48\textwidth}
\begin{tikzpicture}
\begin{axis}[
    width=\textwidth,
    height=0.75\textwidth,
    xlabel={Epoch},
    ylabel={Test Accuracy (\%)},
    xmin=0, xmax=200,
    ymin=10, ymax=67,
    grid=major,
    grid style={gray!30},
    legend style={at={(0.98,0.02)}, anchor=south east, font=\tiny, draw=none, fill=white, fill opacity=0.8},
    every axis plot/.append style={thick},
]
\addplot[color=transformer, thick] table[x=epoch, y=test/acc, col sep=comma]
    {../report_data/transformer_D100_H4.csv};
\addlegendentry{Transf. H=4 (65.2\%)}
\addplot[color=transformer-pm, thick, dashed] table[x=epoch, y=test/acc, col sep=comma]
    {../report_data/transformer_L10_D100_H200_B64_warmup20.csv};
\addlegendentry{Transf. H=1 (63.5\%)}
\addplot[color=deltanet-ffn, thick] table[x=epoch, y=test/acc, col sep=comma]
    {../report_data/deltanet_D104_H4_ffn.csv};
\addlegendentry{DN+FFN H=4 (63.6\%)}
\addplot[color=deltanet-pm, thick, dashed] table[x=epoch, y=test/acc, col sep=comma]
    {../report_data/deltanet_L10_D104_H200_B64_ffn_warmup20_paramatch.csv};
\addlegendentry{DN+FFN H=1 (62.8\%)}
\addplot[color=gfr-ffn, thick] table[x=epoch, y=test/acc, col sep=comma]
    {../report_data/gfr_K1_D115_ffn.csv};
\addlegendentry{GFR+FFN (58.9\%)}
\addplot[color=apn-ffn, thick] table[x=epoch, y=test/acc, col sep=comma]
    {../report_data/apn_L10_D115_H200_B64_ffn_warmup20_paramatch.csv};
\addlegendentry{APN+FFN (56.8\%)}
\end{axis}
\end{tikzpicture}
\caption{Test accuracy}
\end{subfigure}
\hfill
\begin{subfigure}[b]{0.48\textwidth}
\begin{tikzpicture}
\begin{axis}[
    width=\textwidth,
    height=0.75\textwidth,
    xlabel={Epoch},
    ylabel={Test Loss},
    xmin=0, xmax=200,
    ymin=0.5, ymax=6.0,
    grid=major,
    grid style={gray!30},
    legend style={at={(0.02,0.98)}, anchor=north west, font=\tiny, draw=none, fill=white, fill opacity=0.8},
    every axis plot/.append style={thick},
]
\addplot[color=transformer, thick] table[x=epoch, y=test/loss, col sep=comma]
    {../report_data/transformer_D100_H4.csv};
\addlegendentry{Transf. H=4}
\addplot[color=transformer-pm, thick, dashed] table[x=epoch, y=test/loss, col sep=comma]
    {../report_data/transformer_L10_D100_H200_B64_warmup20.csv};
\addlegendentry{Transf. H=1}
\addplot[color=deltanet-ffn, thick] table[x=epoch, y=test/loss, col sep=comma]
    {../report_data/deltanet_D104_H4_ffn.csv};
\addlegendentry{DN+FFN H=4}
\addplot[color=deltanet-pm, thick, dashed] table[x=epoch, y=test/loss, col sep=comma]
    {../report_data/deltanet_L10_D104_H200_B64_ffn_warmup20_paramatch.csv};
\addlegendentry{DN+FFN H=1}
\addplot[color=gfr-ffn, thick] table[x=epoch, y=test/loss, col sep=comma]
    {../report_data/gfr_K1_D115_ffn.csv};
\addlegendentry{GFR+FFN}
\addplot[color=apn-ffn, thick] table[x=epoch, y=test/loss, col sep=comma]
    {../report_data/apn_L10_D115_H200_B64_ffn_warmup20_paramatch.csv};
\addlegendentry{APN+FFN}
\end{axis}
\end{tikzpicture}
\caption{Test loss}
\end{subfigure}
\caption{$\approx$1.2M parameter group. All models except APN+FFN exhibit strong loss--accuracy divergence after reaching 100\% train accuracy. APN+FFN reaches 100\% train accuracy but shows a milder loss increase, consistent with its lower confidence overall. The Transformer and DeltaNet+FFN both gain from $H{=}4$ multi-head attention.}
\label{fig:app_1.2M}
\end{figure}

\subsection{Group 2: $\approx$103K Parameter Models (No FFN)}

Models in this group are capacity-limited and generally do not reach 100\% train accuracy. Overfitting patterns vary dramatically by architecture.

\begin{figure}[h]
\centering
\begin{subfigure}[b]{0.48\textwidth}
\begin{tikzpicture}
\begin{axis}[
    width=\textwidth,
    height=0.75\textwidth,
    xlabel={Epoch},
    ylabel={Test Accuracy (\%)},
    xmin=0, xmax=200,
    ymin=30, ymax=66,
    grid=major,
    grid style={gray!30},
    legend style={at={(0.98,0.02)}, anchor=south east, font=\tiny, draw=none, fill=white, fill opacity=0.8},
    every axis plot/.append style={thick},
]
\addplot[color=apan-rmsnorm, thick] table[x=epoch, y=test/acc, col sep=comma]
    {../report_data/apan_L10_D100_rmsnorm_noFFN.csv};
\addlegendentry{APAN RMSNorm (64.2\%)}
\addplot[color=apan-tanh, thick] table[x=epoch, y=test/acc, col sep=comma]
    {../report_data/apan_L10_D100_tanh_noFFN.csv};
\addlegendentry{APAN tanh (56.4\%)}
\addplot[color=apn-none, thick] table[x=epoch, y=test/acc, col sep=comma]
    {../report_data/apn_L10_D100_H200_B64_act-none_warmup20.csv};
\addlegendentry{APN none (58.3\%)}
\addplot[color=apn-tanh, thick] table[x=epoch, y=test/acc, col sep=comma]
    {../report_data/apn_L10_D100_H200_B64_act-tanh_better_init_lam_eta.csv};
\addlegendentry{APN tanh (53.2\%)}
\addplot[color=deltanet-pm, thick] table[x=epoch, y=test/acc, col sep=comma]
    {../report_data/deltanet_L10_D58_H200_B64.csv};
\addlegendentry{DeltaNet D=58 (60.1\%)}
\addplot[color=transformer-pm, thick] table[x=epoch, y=test/acc, col sep=comma]
    {../report_data/transformer_L10_D29_H200_B64_warmup20_paramatch.csv};
\addlegendentry{Transf. D=29 (59.8\%)}
\addplot[color=gfr-k4, thick] table[x=epoch, y=test/acc, col sep=comma]
    {../report_data/gfr_K4_D100_perslot.csv};
\addlegendentry{GFR K=4 (56.0\%)}
\addplot[color=gfr-k1, thick] table[x=epoch, y=test/acc, col sep=comma]
    {../report_data/gfr_K1_D100.csv};
\addlegendentry{GFR K=1 (53.6\%)}
\end{axis}
\end{tikzpicture}
\caption{Test accuracy}
\end{subfigure}
\hfill
\begin{subfigure}[b]{0.48\textwidth}
\begin{tikzpicture}
\begin{axis}[
    width=\textwidth,
    height=0.75\textwidth,
    xlabel={Epoch},
    ylabel={Test Loss},
    xmin=0, xmax=200,
    ymin=1.0, ymax=2.5,
    grid=major,
    grid style={gray!30},
    legend style={at={(0.98,0.98)}, anchor=north east, font=\tiny, draw=none, fill=white, fill opacity=0.8},
    every axis plot/.append style={thick},
]
\addplot[color=apn-none, thick] table[x=epoch, y=test/loss, col sep=comma]
    {../report_data/apn_L10_D100_H200_B64_act-none_warmup20.csv};
\addlegendentry{APN none}
\addplot[color=apn-tanh, thick] table[x=epoch, y=test/loss, col sep=comma]
    {../report_data/apn_L10_D100_H200_B64_act-tanh_better_init_lam_eta.csv};
\addlegendentry{APN tanh}
\addplot[color=deltanet-pm, thick] table[x=epoch, y=test/loss, col sep=comma]
    {../report_data/deltanet_L10_D58_H200_B64.csv};
\addlegendentry{DeltaNet D=58}
\addplot[color=transformer-pm, thick] table[x=epoch, y=test/loss, col sep=comma]
    {../report_data/transformer_L10_D29_H200_B64_warmup20_paramatch.csv};
\addlegendentry{Transf. D=29}
\addplot[color=gfr-k4, thick] table[x=epoch, y=test/loss, col sep=comma]
    {../report_data/gfr_K4_D100_perslot.csv};
\addlegendentry{GFR K=4}
\addplot[color=gfr-k1, thick] table[x=epoch, y=test/loss, col sep=comma]
    {../report_data/gfr_K1_D100.csv};
\addlegendentry{GFR K=1}
\end{axis}
\end{tikzpicture}
\caption{Test loss}
\end{subfigure}
\caption{$\approx$103K parameter group (no FFN). DeltaNet D=58 and Transformer D=29 both peak early and decline significantly (4.7 and 3.9 point drops respectively), despite reaching 100\% train accuracy. APN and GFR models show steady improvement with minimal overfitting, as they cannot fully memorize the training set. Test loss for APN/GFR stabilizes around 1.2--1.3, while DeltaNet/Transformer test loss increases after their peaks.}
\label{fig:app_103K}
\end{figure}

\subsection{Group 3: GFR Configurations}

\begin{figure}[h]
\centering
\begin{subfigure}[b]{0.48\textwidth}
\begin{tikzpicture}
\begin{axis}[
    width=\textwidth,
    height=0.75\textwidth,
    xlabel={Epoch},
    ylabel={Test Accuracy (\%)},
    xmin=0, xmax=200,
    ymin=30, ymax=60,
    grid=major,
    grid style={gray!30},
    legend style={at={(0.98,0.02)}, anchor=south east, font=\tiny, draw=none, fill=white, fill opacity=0.8},
    every axis plot/.append style={thick},
]
\addplot[color=gfr-ffn, thick] table[x=epoch, y=test/acc, col sep=comma]
    {../report_data/gfr_K1_D115_ffn.csv};
\addlegendentry{K=1+FFN D=115 (58.9\%)}
\addplot[color=gfr-k4, thick] table[x=epoch, y=test/acc, col sep=comma]
    {../report_data/gfr_K4_D100_perslot.csv};
\addlegendentry{K=4 per-slot (56.0\%)}
\addplot[color=gfr-k4, thick, dashed] table[x=epoch, y=test/acc, col sep=comma]
    {../report_data/gfr_K4_D100_perneuron.csv};
\addlegendentry{K=4 per-neuron (56.2\%)}
\addplot[color=gfr-k1, thick] table[x=epoch, y=test/acc, col sep=comma]
    {../report_data/gfr_K1_D100.csv};
\addlegendentry{K=1 per-slot (53.6\%)}
\addplot[color=gfr-k1, thick, dashed] table[x=epoch, y=test/acc, col sep=comma]
    {../report_data/gfr_K1_D100_perneuron.csv};
\addlegendentry{K=1 per-neuron (52.4\%)}
\addplot[color=gfr-k1, thick, dotted] table[x=epoch, y=test/acc, col sep=comma]
    {../report_data/gfr_K1_D34_ffn_paramatch.csv};
\addlegendentry{K=1+FFN D=34 (57.6\%\textrightarrow 51.9\%)}
\end{axis}
\end{tikzpicture}
\caption{Test accuracy}
\end{subfigure}
\hfill
\begin{subfigure}[b]{0.48\textwidth}
\begin{tikzpicture}
\begin{axis}[
    width=\textwidth,
    height=0.75\textwidth,
    xlabel={Epoch},
    ylabel={Test Loss},
    xmin=0, xmax=200,
    ymin=0.5, ymax=5.5,
    grid=major,
    grid style={gray!30},
    legend style={at={(0.02,0.98)}, anchor=north west, font=\tiny, draw=none, fill=white, fill opacity=0.8},
    every axis plot/.append style={thick},
]
\addplot[color=gfr-ffn, thick] table[x=epoch, y=test/loss, col sep=comma]
    {../report_data/gfr_K1_D115_ffn.csv};
\addlegendentry{K=1+FFN D=115}
\addplot[color=gfr-k4, thick] table[x=epoch, y=test/loss, col sep=comma]
    {../report_data/gfr_K4_D100_perslot.csv};
\addlegendentry{K=4 per-slot}
\addplot[color=gfr-k1, thick] table[x=epoch, y=test/loss, col sep=comma]
    {../report_data/gfr_K1_D100.csv};
\addlegendentry{K=1 per-slot}
\addplot[color=gfr-k1, thick, dotted] table[x=epoch, y=test/loss, col sep=comma]
    {../report_data/gfr_K1_D34_ffn_paramatch.csv};
\addlegendentry{K=1+FFN D=34}
\end{axis}
\end{tikzpicture}
\caption{Test loss}
\end{subfigure}
\caption{All GFR configurations. Key observations: (1) $K{=}4$ gains $\approx$2.4 points over $K{=}1$. (2) Per-neuron and per-slot perform similarly for $K{=}4$, but per-neuron is \emph{worse} for $K{=}1$. (3) GFR $K{=}1$+FFN at $D{=}34$ (103K params) peaks at 57.6\% (epoch 39) but overfits dramatically to 51.9\%---the FFN memorizes with the small-D recurrence. (4) GFR $K{=}1$+FFN at $D{=}115$ (1.2M) shows the loss--accuracy divergence pattern (test loss $\to$ 5.3, acc stable at 58.9\%).}
\label{fig:app_gfr}
\end{figure}

\subsection{Group 4: Width-Matched DeltaNet ($D{=}100$)}

\begin{figure}[h]
\centering
\begin{tikzpicture}
\begin{axis}[
    width=0.7\textwidth,
    height=0.45\textwidth,
    xlabel={Epoch},
    ylabel={Test Accuracy (\%)},
    xmin=0, xmax=200,
    ymin=35, ymax=64,
    grid=major,
    grid style={gray!30},
    legend style={at={(0.98,0.02)}, anchor=south east, font=\footnotesize, draw=none, fill=white, fill opacity=0.8},
    every axis plot/.append style={thick},
]
\addplot[color=deltanet, thick] table[x=epoch, y=test/acc, col sep=comma]
    {../report_data/deltanet_L10_D100_H200_B64.csv};
\addlegendentry{DeltaNet D=100 no warmup (62.3\%)}
\addplot[color=deltanet-pm, thick, dashed] table[x=epoch, y=test/acc, col sep=comma]
    {../report_data/deltanet_L10_D100_H200_B64_warmup20.csv};
\addlegendentry{DeltaNet D=100 warmup20 (59.3\%)}
\addplot[color=apn-none, thick] table[x=epoch, y=test/acc, col sep=comma]
    {../report_data/apn_L10_D100_H200_B64_act-none_warmup20.csv};
\addlegendentry{APN D=100 (58.3\%, 3$\times$ fewer params)}
\end{axis}
\end{tikzpicture}
\caption{Width-matched comparison at $D{=}100$. DeltaNet (305K params) peaks at 62.3\% (epoch 11) but declines to 60.8\%. Warmup \emph{hurts} DeltaNet by 3 points. APN (103K params, $3\times$ fewer) achieves 58.3\% with steady convergence and no overfitting.}
\label{fig:app_width}
\end{figure}

\subsection{Run Configuration Details}

\begin{table}[h]
\centering
\caption{Complete configuration details for all 26 runs. All use $L{=}10$ layers, batch size 64, AdamW with lr=$10^{-3}$, cosine schedule.}
\label{tab:app_configs}
\resizebox{\textwidth}{!}{%
\begin{tabular}{llcccccl}
\toprule
\textbf{Run Name} & \textbf{Model} & \textbf{$D$} & \textbf{$H$} & \textbf{FFN} & \textbf{Params} & \textbf{Warmup} & \textbf{Notes} \\
\midrule
\texttt{transformer\_D100\_H4} & Transformer & 100 & 4 & Yes & 1,205,610 & 20 & Best overall (65.2\%) \\
\texttt{transformer\_L10\_D100\_\ldots\_warmup20} & Transformer & 100 & 1 & Yes & 1,205,610 & 20 & H=1 baseline \\
\texttt{transformer\_L10\_D29\_\ldots\_paramatch} & Transformer & 29 & 1 & Yes & 102,554 & 20 & Param-matched \\
\texttt{deltanet\_D104\_H4\_ffn} & DeltaNet+FFN & 104 & 4 & Yes & 1,199,754 & 20 & Multi-head DeltaNet \\
\texttt{deltanet\_L10\_D104\_\ldots\_paramatch} & DeltaNet+FFN & 104 & 1 & Yes & 1,196,634 & 20 & 1.2M param-match \\
\texttt{deltanet\_L10\_D100\_\ldots} & DeltaNet & 100 & 1 & No & 304,610 & 0 & Width-matched \\
\texttt{deltanet\_L10\_D100\_\ldots\_warmup20} & DeltaNet & 100 & 1 & No & 304,610 & 20 & Warmup hurts \\
\texttt{deltanet\_L10\_D58\_\ldots} & DeltaNet & 58 & 1 & No & 103,598 & 0 & Param-matched \\
\texttt{apn\_L10\_D100\_\ldots\_act-none\_warmup20} & APN & 100 & 1 & No & 103,630 & 20 & Best APN (58.3\%) \\
\texttt{apn\_L10\_D100\_\ldots\_act-none} & APN & 100 & 1 & No & 103,630 & 0 & No warmup (58.1\%) \\
\texttt{apn\_L10\_D115\_\ldots\_ffn\_\ldots\_paramatch} & APN+FFN & 115 & 1 & Yes & 1,196,720 & 20 & FFN hurts (56.8\%) \\
\texttt{apn\_D116\_H4\_ffn\_perhead} & APN+FFN & 116 & 4 & Yes & 1,217,626 & 20 & NaN at epoch 10 \\
\texttt{apn\_D100\_H4\_perhead} & APN & 100 & 4 & No & 103,690 & 20 & NaN from epoch 1 \\
\texttt{apn\_L10\_D100\_\ldots\_act-tanh\_better\_init} & APN & 100 & 1 & No & 103,630 & 0 & $\lambda{=}0.999,\eta{=}1.0$ \\
\texttt{apn\_L10\_D100\_\ldots\_act-tanh} & APN & 100 & 1 & No & 103,630 & 0 & Default init \\
\texttt{apn\_freeze-lam\_\ldots} & APN & 100 & 1 & No & 103,630 & 0 & Frozen $\lambda$ \\
\texttt{apn\_L10\_D173\_\ldots\_act-tanh} & APN & 173 & 1 & No & 305,548 & 0 & Tile-inefficient \\
\texttt{apn\_L10\_D100\_H200\_B1024} & APN & 100 & 1 & No & 103,630 & 0 & Large batch (800ep) \\
\texttt{gfr\_K1\_D100} & GFR $K{=}1$ & 100 & 1 & No & 103,630 & 20 & Simplest GFR \\
\texttt{gfr\_K1\_D100\_perneuron} & GFR $K{=}1$ & 100 & 1 & No & 105,610 & 20 & Per-neuron $\lambda/\eta$ \\
\texttt{gfr\_K4\_D100\_perslot} & GFR $K{=}4$ & 100 & 1 & No & 103,690 & 20 & 4 timescales \\
\texttt{gfr\_K4\_D100\_perneuron} & GFR $K{=}4$ & 100 & 1 & No & 111,610 & 20 & Per-neuron (4 slots) \\
\texttt{gfr\_K1\_D115\_ffn} & GFR+FFN & 115 & 1 & Yes & 1,196,720 & 20 & 1.2M param-match \\
\texttt{gfr\_K1\_D34\_ffn\_paramatch} & GFR+FFN & 34 & 1 & Yes & 105,974 & 20 & 103K param-match \\
\bottomrule
\end{tabular}%
}
\end{table}

\end{document}
